Bike Share data reveiw by Tyler Eller
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
options(scipen=999)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
cites <- read.csv("cites.csv")
ny_chi_collection <- read.csv('ny_chi_collection.csv')


#make sure the data looks as expected. 
head(cites,5)

```

# Univariate Plots Section


Introduction: This data reviews data provided by Motivate about the bike-sharing 
system in New York, Washington D.C, and Chicago. We will mainly be focusing on 
the customer in this data set. 

let's take a look at the structure of the data. 
```{r echo = FALSE}
str(cites)

```


### Gender Breakdown 
```{r echo=FALSE,}
#let's start by looking at the gender distribution 
total <- sort(table(ny_chi_collection$Gender))
na.omit(total)

```

### Gender Breakdown by Percentage
```{r echo=FALSE,}

total <- sort(table(ny_chi_collection$Gender))
round((total/length(ny_chi_collection$Gender)*100), digits =2 )

```


```{r echo = FALSE}

ggplot(aes(x = Gender), data = na.omit(ny_chi_collection)) +
    geom_bar(position = 'dodge', color="black") +
    ggtitle('Gender Breakdown of Riders') +
    scale_x_discrete(labels = c('Female', 'Male')) +
    labs(y = 'Number of Riders', x = 'Gender')

```

You can see that there are way more male riders then female riders. More double
almost. Yet, a count alone can be hard to realy draw uppon. Let's take a look
at precetages. 


```{r echo = FALSE}

#First we need to transform the data to have the percentage. 
gender_perc <- ny_chi_collection %>%
  group_by(Gender) %>%
  count() %>%
  ungroup() %>%
  mutate(perc = `n`/ sum(`n`)) %>%
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

ggplot(gender_perc, aes(x="", y=perc, fill = Gender))+
  geom_col()+
  labs(title = "Gender of Riders(Percentage) ", x = "",  y ="")+
  geom_text(aes(label = labels), position = position_stack(vjust = 0.5))+
  coord_polar(theta = "y")
```


We can now see truly how many more males there are compared to females or 
people who did not proclaim male or female. Males make up 64.2% of the over all
user base. 

### Age

What is the most common age of riders?
Is there a pattern of usage?

```{r echo = FALSE}
#Create a function to make age tables
age_table_creation <- function(data){
    age_table <- table(data$Age)
    age_table <- sort(age_table, decreasing=TRUE)
    age_table
}

#Let's look at the age breakdown of cites. 
age_table_creation(cites)
```
Here you can see that the top five users are in their early 30's and that a 
large part of their user base is in their early 30's. Also, I'm assuming that a 
few of their outlets are incorrect or fabrication, such the people at 134.

```{r echo = FALSE}
ggplot(aes(x= Age), data = cites) +
    geom_bar(position = 'dodge', color="black") +
    ggtitle('Ages of Rides') +
    labs(y = 'Number of Riders', x = 'Age ') 
```

The above graph seems to be stretched out by outlines. 
Lets scale down to 20-60 to see if the data is better. 


```{r echo = FALSE}
ggplot(aes(x= Age), data = cites) +
    geom_histogram(position = 'dodge', color="black") +
    scale_x_continuous(limits = c(20,60))+
    ggtitle('Age of Riders') +
    labs(y = 'Numbers of Riders', x = 'Age') 
```

Here you can see that the highest amount of users are in their early 30's
I do find the random spikes in their 50's and 60's to be interesting and would 
love to delve into what causes them. 


### User types 

let's take a quick look at the diffrent type of users. 
```{r echo = FALSE}
cites %>%
count(User.Type)

```


```{r echo = FALSE}
ggplot(aes(x= User.Type), data = cites) +
    geom_bar(position = 'dodge', color="black") +
    ggtitle('Count of User Types ') +
    labs(y = 'Number of Riders', x="User Types") 


```

Just as with Gender, there is a definitve majority with Subscriber. 

But just as with the Gender, a count is hard to tell the true meaning of the 
size diffrence, so let's lookk at the percentage. 

```{r echo = FALSE}

#First we need to transform the data to have the percentage. 
user_perc <- cites %>%
  group_by(User.Type) %>%
  count() %>%
  ungroup() %>%
  mutate(perc = `n`/ sum(`n`)) %>%
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

ggplot(user_perc, aes(x="", y=perc, fill = User.Type))+
  geom_col()+
  geom_text(aes(label = labels), position = position_stack(vjust = 0.3))+
  coord_polar(theta = "y")+
  ggtitle('User Types (Percentages) ')+
  labs(x="", y = "")
  
```

Subscriber is by far the most common type of user with 81% (rounded up).


```{r echo = FALSE}
summary(cites$Trip.Duration)
```

The average durations of a trip is 17 min. 

# Univariate Analysis 

### What is the structure of the data?
There are 900000 observations of customers with ten variables.
Observations:
- Most of the customers are in the age range of 30-34. 
- Bike share seems to be way more popular with male customers. 
- there is an equal amount of observations across the cities.
- The most common starting destination is 1 Ave & E 16 St as the most common
  Endstation. Could this be a sign that most trips are round trips? 



### What is/are the main feature(s) of interest in your dataset?
I found the age and gender data to be interesting. This data could be analyzed. 
This
data could help better understand Motivate customer base.


### What other features in the dataset do you think will help support your \
###investigation into your feature(s) of interest?

I think delving into the users' stats by the city could be interesting. If one 
city 
has a higher customer count that is female you could explore why that is and
see if you could increase the number of female users in other cities. 


### Did you create any new variables from existing variables in the dataset?
Yes, I made the Age category as it was easier to read than just looking at
the birth_year type. 

### Of the features you investigated, were there any unusual distributions? \
###Did you perform any operations on the data to tidy, adjust, or change the form \
###of the data? If so, why did you do this?

No, this data was pretty clean and did not need any wrangling. 
wrangling


# Bivariate Plots Section

Earlier, we took a look at genders for all cites, now let's take a look at
at Genders broken down by city

```{r echo=FALSE, Bivariate_Plots}

#let's now take a look at gender broken down by cities.
ggplot(aes(x = Gender, fill = city), data = na.omit(ny_chi_collection)) +
    geom_bar(position = 'dodge', color="black") +
    ggtitle('Gender by City') +
    scale_x_discrete(labels = c('Female', 'Male')) +
    labs(y = 'Number of Riders ', x = 'Gender') +
    scale_fill_manual("Cites", values = c("chi" = "Green", "ny" = "Blue"))

```

We are only able to see Chicago and New York as Washington does not provided
gender data. The service seems more popular in New York then in Chicago, for
both genders. 

Lets look at gender and age. 

```{r echo = FALSE}
ggplot(aes(x= Age, fill = Gender), data = ny_chi_collection) +
    geom_bar(position = 'dodge', color="black") +
    scale_x_continuous(limits = c(20 ,60))+
    ggtitle('Gender By Age Groups') +
    labs(y = 'Number of Riders', x = 'Age Groups') +
    scale_fill_manual("Gender Key", values = c("Male" = "Green", "Female" = "Blue" ))
```

This chart shows the gap between male and email users. Female users in
their early 20's seem to be the closest to their male counterpart. If I was
designing a business strategy proposal around this data, I would suggest a 
campaign to get more young female riders as they seem the most interested and 
have the opportunity to stay the longest with the service. 

As Males are so much more than females, it's hard to tell they are True count.
Let's dive down into the data and look at females only.

```{r echo = FALSE}

female <- subset(ny_chi_collection, Gender == "Female",select = X:city)

ggplot(aes(x= Age, fill = Gender), data = female) +
    geom_bar(position = 'dodge', color="black")+
    scale_x_continuous(limits = c(20 ,60))+
    ggtitle('Riders Ages (Female only) ') +
    labs(y = 'Number of Riders', x = 'Age') 
```

While the number of male customers outweighs the number of female users, I
find it interesting that they have the same shape. I wonder if 
the bike share service does have many under 25 users, and most are over the age 
of 30 due to income overage. It may be that at 30, people start having enough
income to use on extra survives for leisure items. Another data point I'd love to 
delve into is the reason for use, looking at lessor or transportation.  

Just to compaire lets look at the male only table. 

```{r echo = FALSE}

Male <- subset(ny_chi_collection, Gender == "Male",select = X:city)

ggplot(aes(x= Age, fill = Gender), data = Male) +
    geom_bar(position = 'dodge', color="black", fill = "Blue")+
    scale_x_continuous(limits = c(20 ,60))+
    ggtitle('Riders Ages (Male only) ') +
    labs(y = 'Number of Riders', x = 'Age') 
```

Males in their early 30's is the largest demographic for this service. 

```{r echo = FALSE}
ggplot(ny_chi_collection, aes(x =Gender, y= Age))+
geom_boxplot(notch = TRUE, outlier.colour = "red")+
#scale_y_continuous(limits = c(20,100))+  
labs(title="Age And Gender Boxplot", x = "Gender")
```

The above box plot aluded to an interesting point the other parts didn't that 
Males, seem to have just a slightly older user base going into their 50's. 

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x= User.Type, fill = city), data = cites) +
    geom_bar(position = 'dodge', color="black") +
    ggtitle('User Types by City') +
    labs(y = 'Rider Count', x = 'User Types') +
    scale_fill_manual("Cites", values = c("chi" = "Blue","ny" = "Green", "wash" = "Gray"))
```

Washington has the most customer type of users. While New York has a strong 
subscriber.

```{r echo = FALSE}
ggplot(aes(x=Age , y = Trip.Duration), data = cites)+
  geom_point()+
  #scale_x_continuous(limits = c(20,100))+
  #scale_y_continuous(limits = c(60,50000))+
  labs(title="Trip Duration Compard to Age", y = "Trip Duration (in minutes)")
```

Due to the outlines this graph does not really help us learn anything. Let's 
zoom in a little to see if we get a better understanding of the data. 

```{r echo=FALSE}
ggplot(aes(x=Age , y = Trip.Duration), data = cites)+
  geom_point(alpha = 0.5)+
  xlim(20,60)+
  ylim(0,700000)+
  labs(title="Trip Duration Compard to Age", y = "Trip Duration (in minutes)")
```

Hmm, this really tells less about the avrage users. Let's zoom in on the Y acess.  


```{r echo=FALSE}
ggplot(aes(x=Age , y = Trip.Duration), data = cites)+
  geom_jitter(alpha = 0.5)+
  xlim(20,60)+
  ylim(0,10000)+
  labs(title="Trip Duration Compard to Age", y = "Trip Duration (in minutes)")


```



```{r echo=FALSE}
ggplot(aes(x= User.Type, fill = city), data = cites) +
    geom_bar(position = 'dodge', color="black") +
    ggtitle('User Types by Cities') +
    labs(y = 'Rider Count', x = 'User Types') +
    scale_fill_manual("Cites", values = c("chi" = "Blue","ny" = "Green",
                                          "wash" = "Red"))

```


Note that Washington DC has the highest customer rate. 
This could be interesting to find out why and if you could incress the customer in 
New York, with out a decline on Subscribers. 

# Multivariate Plots Section

Is there any interesting correlation between Age, Trip Duration and Gender?

Let's take a look at the same break out, but this time with a box plot. 

```{r echo=FALSE}
ggplot(aes(x=Age , y = Trip.Duration, color=Gender), data = cites)+
  geom_point()+
  xlim(20,80)+
  ylim(60,1000)+
  labs(title="Trip Duration Vs Gender and Age", y = "Trip Duration (in Minutes)")
```

Age and Gender does not seem to have any change on the trip duration. 

Let's focus and only look at the females to see if there are any noticeable 
patterns 

```{r echo = FALSE}
ggplot(aes(x=Age , y = Trip.Duration), data = female)+
  geom_point()+
  scale_x_continuous(limits = c(20,100))+
  scale_y_continuous(limits = c(60,1000))+
  labs(title="Trip Duration vs Age for Female Riders ", y = "Trip Duration (in Minutes)")
```

No real patterns jump out of the data when I removed the male and NA data.  

Is there any interesting correlation between Age, Trip Duration and Cities?

```{r echo = FALSE}
ggplot(aes(x=Age , y = Trip.Duration, color=city), data = cites)+
  geom_point()+
  scale_x_continuous(limits = c(20,100))+
  scale_y_continuous(limits = c(60,1000))+
  labs(title="Trip Duration Vs Age and City", y = 'Trip Duration (in Minutes)' )
```


Cites don't seem to play a large effect on the durarion of the trip. Yet, 
New Your does have te oldest riders. 

```{r echo = FALSE}
ggplot(aes(x=Age , y = Trip.Duration, color=User.Type), data = cites)+
  geom_point()+
  scale_x_continuous(limits = c(20,100))+
  scale_y_continuous(limits = c(60,1000))+
  labs(title="Trip Duration by Age and  Customer Type", y = "Trip Duration (in Minutes)")
```

The customer type seem mostly to stay in short trips under 500 min. 
Where Subscriber's have the longer trips. This seems to have the largest change
on a riders duration.

# Multivariate Analysis

In this section of code, I was trying to see how gender, age, city, and 
subscriber type to see if we could figure out who the super users are. 
These are the 
outliers of trip distance. 

In the first graph, you can tell that up until their 50's super users are just 
about
as likely to be male and female. And that some of the most extended trips belong 
tofemales. 

The second chart shows that superusers are also most likely to be from New.
York or Chicago. Though a little is more likely to be someone from New York. 

In the third chart, unexpectedly supers users are subscribers. 

If I were to go further with this analysis, I would be interested in adding some 
machine learning to be able to use this data to find a clear answer to this supper 
user. Also, I'd love to see more historical data so that I can make predictions 
on changes.As well as run some experimental campaigns and see if it changes the
data at all. 


------

# Final Plots and Summary

### Plot One
```{r echo = FALSE}
ggplot(aes(x= Age, fill = Gender), data = cites) +
    geom_histogram(position = 'dodge', color="black",bins = 20) +
    scale_x_continuous(limits = c(20,60))+
    ggtitle('Gender and Age ') +
    labs(y = 'Number of Riders', x = 'Age') +
    scale_fill_manual("Gender Key", values = c("NA" = "Red","Male" = "Green", "Female" = "Blue" ))
```


### Description One
This graph show gender vers age. While both graphs have the same shape the weight
of having males be 64% of the users is really shown in this graph. 

### Plot Two
```{r echo = FALSE}
ggplot(aes(x=Age , y = Trip.Duration, color=city), data = cites)+
  geom_point()+
  scale_x_continuous(limits = c(20,100))+
  scale_y_continuous(limits = c(60,1000))+
  labs(title="Trip Duration Vs Age and City", y = 'Trip Duration (in Minutes)' )
```

### Description Two
This graph shows the diffrence between age and gender in the affects of long 
distance rides. As you can see while there are many more male riders, the 
distrbution of longdistance riders is still preety even. 

### Plot Three
```{r echo = FALSE}
ggplot(aes(x=Age , y = Trip.Duration, color=User.Type), data = cites)+
  geom_point()+
  scale_x_continuous(limits = c(20,100))+
  scale_y_continuous(limits = c(60,1000))+
  labs(title="Trip Duration by Age and  Customer Type", y = "Trip Duration (in Minutes)")
```

### Description Three
Here is a plot showing the age and trip Duration compared to User types. 
Looks like most of the long distance riders already have subscribtions.

------

# Reflection
In the first plot, we genuinely see the difference in male/female riders and the 
shape of age. The vast amount of customers that are male in the age of 
30-35 shaped all other data points that we looked at in this analysis.


It was in final polt three in the analyses of the User types, Age and Trip 
The duration that I discovered something interesting, even though males outweigh 
females, many of super users are female. With the younger generation of 20-25 
females using the service nearly as much as males and many log distant riders
being female, I felt that there could be a massive growth area.

This data set was clean and promising at first. I think the biggest struggle
I came to was wishing I had more data. The Washington D.C. data had no gender 
data, which left me feeling like I was missing a large piece of my puzzle. 
Also, the data set did not tell me the show was repeat customers. So while, in 
final chart # 3 it seems like there is a lot of long distant riders they could 
be a lot of repeat customers.


